Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-28 paper.tex
Move methods to end
Re-break long lines

\subsection{Overview}
Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced
in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
These copying paths
will mostly consist of a new sample sequence copying from a node in the ARG
with a small number of mutations, and often many samples
from a daily batch will copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
This greedy update strategy inevitably lead to unparsimonious
topologies, and the third step is to then increase the
overall parsimony of the inferred ARG by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).
The operations are carried out efficiently using the well-established
\texttt{tskit} library~\citep{Kelleher2018-xc,Ralph2020-efficiently},
and the resulting structure can be conveniently and interactively
analysed using the many features provided by \texttt{tskit}.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny} Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021) and an “all-time” global NextStrain tree (sampled to January 21, 2023). The trees were pruned down to the samples present in both the trees. Light grey lines match the corresponding samples between the two trees; black circles indicate identical sample partitions between the two trees. Terminal branches are colour-coded according to the Pango lineage status assigned to the tip samples; no Pango recombinants are shown, as none are shared between the trees. The tanglegram was generated with the aid of the Neighbor-Net algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5 \citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent cophylogeny for the Long ARG.}
\end{figure}

@@

\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx}
\caption{\label{fig:pango-single-origin-graph}
Single origin Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with imputed Pango status. (B) Details for the single-origin Pango X lineages in the long ARG. (C) XAG origins in the long ARG.}
\end{figure}

@@

% Interestingly, XC is inferred to be single origin in the Wide ARG but not in the long ARG. See lineage subgraphs in supplemental figure Y. This suggests that insufficient sampling may be the explanation for many of these multiple origin events.

@@

\subsection{Novel recombinants}
Next, we examine recombination nodes in the long ARG (sampled to mid-2022) that represent previously unnamed putative recombinant sequences, which do not have a Pango recombinant designation. For this proof-of-concept study, we have arbitrarily picked 12 recombination nodes that seem plausible (Supplementary Table 4). All these nodes involve Omicron subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were inserted into the long ARG on or after January 8, 2022; (2) have at least 10 descendent samples; and (3) have no mutational differences (including immediate reversions) from their parent nodes.

TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006 in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.

@@

>>>>

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees were pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated with the aid of the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

@@

\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-single-origin-graph} Single origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples
and their ancestral lineages are shown outwards to the nearest sampled viral
genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Details for the single-origin Pango X lineages in the
long ARG. (C) XAG origins in the long ARG.
} \end{figure}

@@

@@

% JK: removing this for now

% \subsection{Novel recombinants}
% Next, we examine recombination nodes in the
% long ARG (sampled to mid-2022) that represent previously unnamed putative
% recombinant sequences, which do not have a Pango recombinant designation. For
% this proof-of-concept study, we have arbitrarily picked 12 recombination nodes
% that seem plausible (Supplementary Table 4). All these nodes involve Omicron
% subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
% inserted into the long ARG on or after January 8, 2022; (2) have at least 10
% descendent samples; and (3) have no mutational differences (including immediate
% reversions) from their parent nodes.

% TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
% in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
% sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.

@@

[TODO update this summary para to avoid duplications with
later sections and to add the appropriate signposts
when they are done(ish)]

Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced
in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
These copying paths
will mostly consist of a new sample sequence copying from a node in the ARG
with a small number of mutations, and often many samples
from a daily batch will copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
This greedy update strategy inevitably lead to unparsimonious
topologies, and the third step is to then increase the
overall parsimony of the inferred ARG by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).
The operations are carried out efficiently using the well-established
\texttt{tskit}
library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit},
and the resulting structure can be conveniently and interactively
analysed using the many features provided by \texttt{tskit}.
